
---
title: "<center>FIT5145: Foundations of Data Science</center>
Name: Bhavna Balakrishnan <br> SID:33954427"
output: html_document
date: "2024-04-26"
---

```{r}
#Load all Necessary Libraries:
#Syntax to install any library is for example; install.packages("plotly")
library(ggplot2)
library(plotly)
library(dplyr)
library(lubridate)
library(tm)
library(tidyr)
library(wordcloud)

```

```{r}
# Read the CSV file into a data frame
ireland_news <- read.csv("ireland_news.csv")
```

## Q1. Sorting Articles by time

```{r}

# Convert publish_date column to standard Date format
ireland_news <- ireland_news %>%
  mutate(publish_date = as.Date(publish_date, format = "%A, %dth of %B, %Y"))

# Exclude rows with NA values in publish_date column
ireland_news <- ireland_news %>%
  filter(!is.na(publish_date))

# Sort the data by publish_date in ascending order
ireland_news_sorted <- ireland_news %>%
  arrange(publish_date)

# Display 5 most recent records
head(ireland_news_sorted, 5)

# Display the last 5 records of the sorted data
tail(ireland_news_sorted, 5)
```
Here we can see that the most recent records are from 2021 and the oldest are from 1996.<br>

## Q2. Unique Headline Categories
```{r}
# Convert headline_category column to lowercase to handle variations
ireland_news$headline_category <- tolower(ireland_news$headline_category)

# Count the number of unique headline_category values
num_unique_categories <- length(unique(ireland_news$headline_category))

# Print the result
cat("Number of unique headline category values:", num_unique_categories, "\n")


#2 a

# Convert headline_text column to character type
ireland_news$headline_text <- as.character(ireland_news$headline_text)

# Define the keywords and year range
keywords <- c("ireland", "irish", "us", "usa")
year_range <- 2000:2024

# Construct the regular expression pattern
pattern <- paste0("(\\b(?:", paste(keywords, collapse = "|"), ")\\b.*\\b(?:", paste(year_range, collapse = "|"), ")\\b)")

# Count the number of matches
num_matches <- sum(grepl(pattern, tolower(ireland_news$headline_text)))

# Print the result
cat("Number of news category articles containing the keywords 'Ireland', 'Irish', 'US', or 'USA' along with year digits from 2000 to 2024 in headline_text:", num_matches, "\n")

```
So we can now determine that there are a total number of 110 records with unique headline category values and 1506 articles containing words "Irish","US" and "USA" between years 2000 and 2024. <br>
## Q.3 Top 10 Monday Headlines
```{r}
# Convert publish_date column to Date format
ireland_news$publish_date <- as.Date(ireland_news$publish_date, format = "%A, %dth of %B, %Y")

# Filter the dataset to include only articles published on Mondays
monday_articles <- ireland_news[format(ireland_news$publish_date, "%A") == "Monday", ]

# Count the number of articles for each headline category
category_counts <- table(monday_articles$headline_category)

# Identify the top 10 headline categories with the largest number of articles
top_10_categories <- head(sort(category_counts, decreasing = TRUE), 10)

# Draw a chart showing the total number of articles for the top 10 headline categories for each year

# Filter the dataset to include only the top 10 headline categories
top_categories_data <- monday_articles[monday_articles$headline_category %in% names(top_10_categories), ]

# Create a data frame with the counts of articles for each headline category and year
category_year_counts <- with(top_categories_data, table(headline_category, format(publish_date, "%Y")))

# Convert the data frame to long format for plotting
category_year_counts_long <- as.data.frame(category_year_counts)
colnames(category_year_counts_long) <- c("headline_category", "year", "count")

# Convert the data to a plotly object
p <- ggplot(category_year_counts_long, aes(x = year, y = count, fill = headline_category)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Total Number of Articles for Top 10 Headline Categories (Mondays)",
       x = "Year", y = "Number of Articles") +
  theme_minimal() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5)) +
  scale_fill_discrete(name = "Headline Category")


# Convert to a plotly object and add tooltips
p <- ggplotly(p, tooltip = c("year", "count"))

# Display the plot
p
```
In year 2000 we see number of articles are the most.The major category across years till 2013 appears to be "news"and "sport".After 2013 it's more soccer focused and ireland news.<br> 
## Q.4 Headline Information and Statistics

```{r}
# Compute the total number of articles for each headline category and news provider
article_counts <- ireland_news %>%
  group_by(headline_category, news_provider) %>%
  summarise(total_articles = n())

# Print the computed total number of articles for each headline category and news provider
print(article_counts)

# Compute and display the statistical information (Min, Max, and Mean) for each provider in a single command
article_counts %>%
  group_by(news_provider) %>%
  summarise(across(total_articles, list(Min = min, Max = max, Mean = mean)))
```
Above is the article breakup and statistics by provider<br>


## Q.5 Article Breakdown and averages

```{r}
# Convert publish_date to a Date object
ireland_news$publish_date <- as.Date(ireland_news$publish_date, format = "%A, %dth of %B, %Y")

# Filter out rows with NA values in publish_date
ireland_news <- na.omit(ireland_news, cols = "publish_date")

# Compute the total number of articles for each headline category, news provider, and day of the week
article_counts <- ireland_news %>%
  group_by(headline_category, news_provider, day_of_week = format(publish_date, "%A")) %>%
  summarise(total_articles = n())

# Compute the average number of articles for each news provider and the day of the week
average_articles <- article_counts %>%
  group_by(news_provider, day_of_week) %>%
  summarise(average_articles = mean(total_articles, na.rm = TRUE))

# Find the day of the week with the highest average number of articles for each provider
top_day <- average_articles %>%
  group_by(news_provider) %>%
  top_n(1, average_articles)

# Print the results in the specified table format
print(top_day)

```
Here we have the average articles by provider.<br>
We observe that Irish Times has the maximum average articles and that Irish independent has the least.<br>

## Q.6 Assigning Periods

```{r}
# Convert publish_date to a Date object
ireland_news$publish_date <- as.Date(ireland_news$publish_date, format = "%A, %dth of %B, %Y")

# Filter out rows with NA values in publish_date
ireland_news <- na.omit(ireland_news, cols = "publish_date")

# Filter the data for the years 2019 and 2020
filtered_data <- ireland_news %>%
  filter(year(publish_date) %in% c(2019, 2020))

# Add a new column named "Period" based on the publish_date value
filtered_data <- filtered_data %>%
  mutate(Period = case_when(
    publish_date >= as.Date("2019-01-01") & publish_date <= as.Date("2019-03-31") ~ "Period 1",
    publish_date >= as.Date("2019-04-01") & publish_date <= as.Date("2019-06-30") ~ "Period 2",
    publish_date >= as.Date("2019-07-01") & publish_date <= as.Date("2019-09-30") ~ "Period 3",
    publish_date >= as.Date("2019-10-01") & publish_date <= as.Date("2019-12-31") ~ "Period 4",
    publish_date >= as.Date("2020-01-01") & publish_date <= as.Date("2020-03-31") ~ "Period 5",
    publish_date >= as.Date("2020-04-01") & publish_date <= as.Date("2020-06-30") ~ "Period 6",
    publish_date >= as.Date("2020-07-01") & publish_date <= as.Date("2020-09-30") ~ "Period 7",
    publish_date >= as.Date("2020-10-01") & publish_date <= as.Date("2020-12-31") ~ "Period 8",
    TRUE ~ "Other"
  ))

# Step 1: Compute the total number of articles for each headline category and period
article_counts <- filtered_data %>%
  group_by(headline_category, Period) %>%
  summarise(total_articles = n(), .groups = "drop")

# Step 2: Identify the top ten headline categories
top_categories <- article_counts %>%
  group_by(headline_category) %>%
  summarise(total_articles = sum(total_articles), .groups = "drop") %>%
  top_n(10, total_articles) %>%
  select(headline_category)

# Step 3: Filter the data to include only the top ten headline categories
filtered_article_counts <- article_counts %>%
  filter(headline_category %in% top_categories$headline_category)

# Step 4: Create a boxplot
ggplot(filtered_article_counts, aes(x = Period, y = total_articles)) +
  geom_boxplot(fill = "skyblue", color = "blue") +
  labs(title = "Total Number of Articles by Period (Top 10 Headline Categories)",
       x = "Period", y = "Total Number of Articles") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
From above, we can conclude that period 4 has the maximum median of articles and that period 8 has the least.<br>

## Q.7 Sample Data and Charts

```{r}
#To ensure reproducibility
set.seed(100)

# Step 1: Sample 1% of the data
sampled_data <- ireland_news %>% sample_frac(0.01)

# Step 2: Perform text pre-processing
corpus <- Corpus(VectorSource(sampled_data$headline_text))
corpus <- tm_map(corpus, content_transformer(tolower))
corpus <- tm_map(corpus, removePunctuation)
corpus <- tm_map(corpus, removeNumbers)
corpus <- tm_map(corpus, removeWords, stopwords("en"))
corpus <- tm_map(corpus, stripWhitespace)

# Step 3: Create a document-term matrix
dtm <- DocumentTermMatrix(corpus)
#print(dtm)

# Step 4: Generate a plot showing the top 10 most frequent words
freq_words <- colSums(as.matrix(dtm))
top_words <- sort(freq_words, decreasing = TRUE)[1:10]

# Create a data frame for plotting
word_freq_df <- data.frame(word = names(top_words), frequency = top_words)

# Plot
ggplot(word_freq_df, aes(x = frequency, y = reorder(word, frequency))) +
  geom_point(size = 3, color = "skyblue") +
  labs(title = "Top 10 Most Frequent Words",
       x = "Frequency", y = "Word") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))

#install.packages("wordcloud")

# Create a word cloud
wordcloud(words = word_freq_df$word, freq = word_freq_df$frequency, 
          min.freq = 1, max.words = 10, random.order = FALSE,
          colors = brewer.pal(8, "Dark2"))

```
The most frequent word in this sample is “New” and least is “court”.<br>

## Q.8 Analysing News Categories and trends over the Years

```{r}

# Define the grouping rules
ireland_news <- ireland_news %>%
  mutate(grouped_category = case_when(
    grepl("sport", headline_category, ignore.case = TRUE) ~ "Sports",
    grepl("business|business_economy", headline_category, ignore.case = TRUE) ~ "Business",
    grepl("culture|culture_books", headline_category, ignore.case = TRUE) ~ "Culture",
    grepl("lifestyle|lifestyle_fashion", headline_category, ignore.case = TRUE) ~ "Fashion.News",
    grepl("news|news_ireland|newses|NEWS", headline_category, ignore.case = TRUE) ~ "News",
    grepl("opinion", headline_category, ignore.case = TRUE) ~ "Opinion",
    TRUE ~ headline_category  # Default to original category if no match
  ))

# Count the frequency of each grouped category
category_counts <- table(ireland_news$grouped_category)

# Plot the pie chart
pie_chart <- ggplot(as.data.frame(category_counts), aes(x = "", y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Pie Chart of Grouped Headline Categories",fill ="Category") +
  theme_void() +
  theme(legend.position = "right")

# Display the pie chart
print(pie_chart)

#now making a line chart

# Define custom color palette
custom_colors <- c("Sports" = "#1f77b4",    # Blue
                   "Business" = "#ff7f0e",  # Orange
                   "Culture" = "#2ca02c",   # Green
                   "Fashion.News" = "#d62728",  # Red
                   "News" = "#9467bd",      # Purple
                   "Opinion" = "#8c564b")   # Brown

# Filter and group the data by category and year
category_year_counts <- ireland_news %>%
  mutate(year = as.numeric(format(as.Date(publish_date), "%Y"))) %>%
  filter(grouped_category %in% c("Sports", "Business", "Culture", "Fashion.News", "News", "Opinion")) %>%
  group_by(grouped_category, year) %>%
  summarise(article_count = n())

# Plot the line graph with custom colors
line_plot <- ggplot(category_year_counts, aes(x = year, y = article_count, color = grouped_category)) +
  geom_line() +
  scale_color_manual(values = custom_colors) +  # Set custom colors
  labs(title = "Number of Articles in Each Category Over the Years",
       x = "Year", y = "Number of Articles",
       color = "Category") +
  theme_minimal()

# Display the line graph
print(line_plot)
```
News takes up the most among the categories and from the trends we see an overall decline across all categories in 2020.The top two here happen to be news and sports<br>