---
title: "Loan Approval Data Visualization"
author: "Bhavna Balakrishnan"
date: "22/05/24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Initialise necessary libraries
library(ggplot2)
library(scales)
library(corrplot)
library(readr)
library(rpart)
library(rpart.plot)
#load dataset
loan_data <- read_csv("data.csv")
```

# Demographic Breakdown of Applicants
## Annual Income
```{r income_histogram, fig.cap="Histogram of Annual Income"}
library(scales)  # Load the scales library for formatting axes

# Plotting the histogram for annual income
ggplot(loan_data, aes(x = income_annum)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  scale_x_continuous(labels = comma) +  # Use comma to format large numbers
  labs(title = "Distribution of Annual Income",
       x = "Annual Income ($)",
       y = "Number of Applicants")
## Education Level
```
### Explanation of the Graph and Its Components

- **Histogram**: This type of graph uses bars to show the frequency of data points within specific intervals. It's particularly useful for understanding the shape of the distribution of numerical data.

- **X-axis (Annual Income)**: The x-axis represents different ranges of annual income in dollars. The use of `scale_x_continuous(labels = comma)` ensures that income values are formatted with commas for easier reading (e.g., $30,000 instead of $3e+4).

- **Y-axis (Number of Applicants)**: The y-axis shows the number of applicants that fall within each income range specified by the bins on the x-axis.

We can therefore intuitively interpret that there is a good number of applicants for all income ranges with the maximum applicants coming from the 5000,000 and least from 10,000,000 income range.


## Education 
```{r education_loan_status_bar_chart, fig.cap="Bar Chart of Education Levels by Loan Status"}
library(ggplot2)

# Plotting the bar chart for education levels categorized by loan status
ggplot(loan_data, aes(x = education, fill = loan_status)) +
  geom_bar(position = "fill") +  # 'fill' stacks the bars and normalizes them
  scale_y_continuous(labels = scales::percent_format()) +  # Converts y-axis to percentage
  labs(title = "Education Level vs. Loan Status",
       x = "Education Level",
       y = "Percentage of Applicants",
       fill = "Loan Status",
       caption = "This bar chart shows the distribution of education levels among loan applicants, segmented by loan status (Approved vs. Rejected). Each segment within the bars represents the proportion of applicants within that education category, allowing for a direct comparison of loan outcomes between graduates and non-graduates.")
```

### Explanation of the Graph and Its Components

- **Stacked Bar Chart**: The chart uses stacked bars where each bar represents an education level (e.g., Graduate, Non-Graduate), and the segments in each bar show the proportion of loan status (Approved, Rejected). The stacking is normalized to fill the height of each bar, which means the heights are comparable across education levels.

- **X-axis (Education Level)**: Indicates the categories of education, such as Graduate and Non-Graduate.

- **Y-axis (Percentage of Applicants)**: Shows the percentage of total applicants within each education category that corresponds to each loan status. This axis is formatted to display percentages for better readability and immediate understanding of proportions.

### Conclusion
There appears to be no significant difference between the education levels and education levels of applicants.

## Number of Dependents

```{r}
ggplot(loan_data, aes(x = factor(no_of_dependents), y = loan_amount)) +
  geom_boxplot(aes(fill = factor(no_of_dependents))) +
  scale_y_continuous(labels = label_comma()) +  # Use comma formatting for the y-axis
  labs(title = "Distribution of Loan Amount by Number of Dependents",
       x = "Number of Dependents",
       y = "Loan Amount ($)",
       fill = "Number of Dependents") +
  theme_minimal()

```
### Interpretation
There is no major difference between the amounts asked and number of dependents but we can observe the quartiles and medians from the boxplot.



# Loan Status
```{r}
loan_status_counts <- table(loan_data$loan_status)
pie(loan_status_counts, labels = names(loan_status_counts), main = "Loan Status Distribution", col = rainbow(length(loan_status_counts)))
```
## Interpretation
We can see that it appears that a lot of applicants have chances of having their loans approved.

#Cibil Score
```{r}
ggplot(loan_data, aes(x = loan_status, y = cibil_score, fill = loan_status)) +
  geom_violin(trim = FALSE) +
  labs(title = "CIBIL Score Distribution by Loan Status",
       x = "Loan Status",
       y = "CIBIL Score") +
  theme_minimal()
```
## Interpretation
We can see that from the graph, higher Cibil Scores have higher approvals.

# Decision Tree

```{r}
# Convert loan_status to a factor 
loan_data$loan_status <- as.factor(loan_data$loan_status)
# Building the decision tree model using 'rpart'
# Predicting loan_status based on income_annum and other predictors
tree_model <- rpart(loan_status ~ income_annum + no_of_dependents + cibil_score + education, data = loan_data, method = "class")

# Display the summary of the tree model
summary(tree_model)

# Plotting the decision tree
rpart.plot(tree_model, extra = 106)  # extra = 106 shows node numbers, splits, and all labels
```
## Interpretation
In line with what we previously discussed, We see that applicants with CIBIL scores less than 550 have higher chance of loan rejection.

# Regression Income v loan amt
``` {r}
# Fitting a linear regression model to analyze the relationship between income and loan amount
model <- lm(loan_amount ~ income_annum, data = loan_data)

# Displaying the summary of the regression model
summary(model)

ggplot(loan_data, aes(x = income_annum, y = loan_amount)) +
  geom_point(aes(color = loan_status), alpha = 0.5) +  # Points colored by loan status
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  # Linear regression line without confidence interval
  scale_x_continuous(labels = label_comma()) +  # Format x-axis labels with commas
  scale_y_continuous(labels = label_comma()) +  # Format y-axis labels with commas
  labs(title = "Regression Analysis: Loan Amount vs. Annual Income",
       x = "Annual Income ($)",
       y = "Loan Amount ($)",
       color = "Loan Status") +
  theme_minimal()
# This plot visualizes the relationship between annual income and loan amount, with a regression line showing the trend.
```
## Interpretation
The regression model analyzing the relationship between annual income and loan amount is highly effective, explaining a significant proportion (86.02%) of the variability in loan amounts. The model’s predictions vary from actual values by an average of about $3,382,000, which may be notable, suggesting further refinement could be beneficial depending on the application’s precision needs. The strong statistical significance of the model indicates that annual income is a crucial predictor of loan amount. Financial institutions can rely on this model for a robust estimation of loan amounts based on applicants’ incomes, though they should also consider other factors potentially influencing loan decisions due to the observed residual error.


















